Search for WIMP Inelastic Scattering Off Xenon Nuclei With Xenon100 Data

### 2017-01-09 Inelastic/analysis.tex
first version to ranny

The inelastic scattering of a WIMP with the nucleus of $^{129}$Xe produces an energy deposit via nuclear recoil with subsequent emission of  
a 39.6~KeV de-exitation photon. 
The largest fraction of the energy released in the event is via electronic recoil due to the emitted photon, this represents an
unusual signature for this kind of detector and brings the possible signal to overlay a phase space region with large backgrounds.
The choosen region of interest for this analysis surrounds the 39.6~KeV xenon line in the cS1-cS2 plane which is further divided into
sub-regions, as shown in Figure~\ref{fig:SR}.

Events are asked, other than falling in the defined region of interest, to fullfill several selection criteria which can be sumarized as:
selection aimed to reduce noise impact including energy (S1) and S2 thresholds, events must be of single scatter nature and fall 
into a predefined fiducial volume. This analysis follows the selection criteria described in detail in \cite{dataAnalysis} for Run-II, 
with only few exceptions. 
%only few modification have been designed specifically for this analysis and discussed below. 
In particular, the selection on S2 width as a function of drift time has been optimized on a sample of events selected from the 39.6~KeV line
and set to a 95\% acceptance on these. Events are required to be single scatter by applying a threshold on the 
second largest S2 peak size,  for this analysis this threshold has been optimized to 160~PE and set constant as function of S2 signal size. 
Finally the chosen fiducial volume corresponds to 34~Kg of liquid xenon.

@@

\subsection{Signal Simulation} 
%description of the simulated signal, few words about cross checks MC matching.
The detector response to inelastic scattering of WIMPs  off $^{129}$Xe nucleus was simulated using an empirical model.
The total deposited energy is divided into two independent contributions: the one related to the 39.6~KeV de-exitation photon and the one  relative to
the simultaneous nuclear recoil of the xenon atom, the number of photons and charge yield detected is simulated separatedly for each contribution
and then added togheter.

The ER induced by the de-exitation photon is simulated assuming a two dimensional normal distribution as its pdf in the cS1-cS2 plane, $f(cS1,cS2)$, 
described (except a normalization factor) in equation~\ref{f:2dgaus}:
\begin{multline}
	f(cS1,cS2)  = exp \Big( -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1 - \mu_{cS1})^2}{\sigma_{cs1}^2} + \\ 
	 \frac{(cS2 - \mu_{cS2})^2}{\sigma_{cs2}^2} - \frac{2\rho(cS1 - \mu_{cs1}) (cS2 - \mu_{cs2})} {\sigma_{cs1}\sigma_{cs2}} \Big] \Big) 
	%f(cS1,cS2) ~ = ~  \frac{1}{2 \pi \sigma_{cs1} \sigma_cs2 \sqrt{1-\rho^{2}} } exp \Big( -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1 - \hat{cS1})^2}{\sigma_{cs1}^2} \Big] \Big)
\label{f:2dgaus}
\end{multline}
where $\mu_{cS1}$ and $\mu_{cS2}$ 
represents the average observed cS1 and cS2 given a 39.6~KeV ER, $\sigma_{cs1}$ and $\sigma_{cs2}$ are the standard deviation in cS1 and cS2 respectively,
while $\rho$ stands for the correlation between cS1 and cS2.
The detector related light yield $L_y$ measured at 39.6~KeV, necessary to evaluate the average number of photon detected ($\mu_{cS1}$),
%with AmBe calibration has been found to have a large systematic uncertainty
%due to the contribution of the nuclear recoil, so the light yield used to describe the number of photon detected 
is obtained as the result of a NEST model~\cite{NEST,Geant1,Geant2} fit to data collected with several lines.  The same model is used to predict the charge yield at
39.6~KeV which is then scaled according to the detector's secondary scintillation gain $Y$, determined from detector response to single electrons~\cite{SingleE}.
Note that the corrected S2 observed by the bottom PMT array is used in this analysis.  
The detector resolution at 39.6~KeV in cS1 and cS2 has been measured to be respectively 15.8\% and 14.7\% and used to extract the standard 
deviations $\sigma_{cs1}$, $\sigma_{cs2}$.  The correlation parameter is assumed to be independent of energy (at least in the considered range) and measured
using the 164~KeV xenon activated line by $^{124}$AmBe calibration data, this line is choosen since allows to disentangle efficiently contribution from nuclear recoil.
The measured correlation is $\rho \, = \, -0.45 \pm 0.10$.

@@

The signal simulation procedure has been validated reproducing the 39.6~KeV xenon line from interaction due to
$^{124}$AmBe source and has been compared to data. For the comparison the proper  $^{124}$AmBe nuclear recoil and acceptances
were simulated. The simulated events were found in agreement  with calibration data within statistical uncertainties.

@@

The main expected background contribution in the region of interest is due material radioactivity, composed mainly by photons  
interacting via compton scattering. Background contribution due to the activation of the xenon 39.6~KeV line from radiogenic neutrons is expected to be negligible.
The background is modeled using data from $^{60}$Co calibration campaign, which are assumed to well represent the background density distribution 
in the cS1-cS2 plane.

@@

Signal model, Energy scale uncertainties.... Uncertainties related to energy scale and more generally to detector response 
are taken into account by relative uncertainties on the measure of $L_y$,$\mathcal{L}_{eff}$,$Y$, $Q_Y$ and $\rho$. Has been shown 
by the simulation that these type of uncertainty affect the total signal yield in the ROI by less than a percent for a wide range of masses,
however they greatly affects the pdf of the signal model in the ROI. For each WIMP mass several signal simulation sample are produced 
warying these parameter respectively $\pm 1 \sigma$, for each region is computed an uncertainty by adding in quadrature the variations
%FIXME: explain this better, the up variation are separated from down variations and added togheter
with respect to nominal of each case, Figure~\ref{fig:unc} is an example of such a systematic uncertainty computation for WIMP mass of 100~GeV.

@@

%FIXME: express this in an understandable way
All the uncertainties discussed here are threated as nuissance parameter in the likelihood, which are contrained by a gaussian distribution.

>>>>

The inelastic scattering of a WIMP with the nucleus of $^{129}$Xe produces an energy deposit via nuclear recoil with subsequent emission of  
a 39.6~keV de-exitation photon. 
The largest fraction of the energy released in the event is via electronic recoil due to the emitted photon, this represents an
unusual signature for this kind of detector and brings the possible signal to overlay a phase space region with large backgrounds.
The choosen region of interest for this analysis surrounds the 39.6~keV xenon line in the cS1-cS2 plane which is further divided into
sub-regions, as shown in Figure~\ref{fig:SR}.

Events are asked, other than falling in the defined region of interest, to fullfill several selection criteria which can be sumarized as:
selection aimed to reduce noise impact including energy (S1) and S2 thresholds, events must be of single scatter nature and fall 
into a predefined fiducial volume. This analysis follows the selection criteria described in detail in \cite{dataAnalysis} for Run-II, 
with only few exceptions. 
%only few modification have been designed specifically for this analysis and discussed below. 
In particular, the selection on S2 width as a function of drift time has been optimized on a sample of events selected from the 39.6~keV line
and set to a 95\% acceptance on these. Events are required to be single scatter by applying a threshold on the 
second largest S2 peak size,  for this analysis this threshold has been optimized to 160~PE and set constant as function of S2 signal size. 
Finally the chosen fiducial volume corresponds to 34~Kg of liquid xenon.

@@

\subsection{Signal Simulation} 
%description of the simulated signal, few words about cross checks MC matching.
The detector response to inelastic scattering of WIMPs  off $^{129}$Xe nucleus was simulated using an empirical model.
The total deposited energy is divided into two independent contributions: the one related to the 39.6~keV de-exitation photon and the one  relative to
the simultaneous nuclear recoil of the xenon atom, the number of photons and charge yield detected is simulated separatedly for each contribution
and then added togheter.

The ER induced by the de-exitation photon is simulated assuming a two dimensional normal distribution as its pdf in the cS1-cS2 plane, $f(cS1,cS2)$, 
described (except a normalization factor) in equation~\ref{f:2dgaus}:
\begin{multline}
	f(cS1,cS2)  = exp \Big( -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1 - \mu_{cS1})^2}{\sigma_{cs1}^2} + \\ 
	 \frac{(cS2 - \mu_{cS2})^2}{\sigma_{cs2}^2} - \frac{2\rho(cS1 - \mu_{cs1}) (cS2 - \mu_{cs2})} {\sigma_{cs1}\sigma_{cs2}} \Big] \Big) 
	%f(cS1,cS2) ~ = ~  \frac{1}{2 \pi \sigma_{cs1} \sigma_cs2 \sqrt{1-\rho^{2}} } exp \Big( -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1 - \hat{cS1})^2}{\sigma_{cs1}^2} \Big] \Big)
\label{f:2dgaus}
\end{multline}
where $\mu_{cS1}$ and $\mu_{cS2}$ 
represents the average observed cS1 and cS2 given a 39.6~keVee , $\sigma_{cs1}$ and $\sigma_{cs2}$ are the standard deviation in cS1 and cS2 respectively,
while $\rho$ stands for the correlation between cS1 and cS2.
The detector related light yield $L_y$ measured at 39.6~keV, necessary to evaluate the average number of photon detected ($\mu_{cS1}$),
%with AmBe calibration has been found to have a large systematic uncertainty
%due to the contribution of the nuclear recoil, so the light yield used to describe the number of photon detected 
is obtained as the result of a NEST model~\cite{NEST,Geant1,Geant2} fit to data collected with several lines.  The same model is used to predict the charge yield at
39.6~keV which is then scaled according to the detector's secondary scintillation gain $Y$, determined from detector response to single electrons~\cite{SingleE}.
Note that the corrected S2 observed by the bottom PMT array is used in this analysis.  
The detector resolution at 39.6~keV in cS1 and cS2 has been measured to be respectively 15.8\% and 14.7\% and used to extract the standard 
deviations $\sigma_{cs1}$, $\sigma_{cs2}$.  The correlation parameter is assumed to be independent of energy (at least in the considered range) and measured
using the 164~keV xenon activated line by $^{124}$AmBe calibration data, this line is choosen since allows to disentangle efficiently contribution from nuclear recoil.
The measured correlation is $\rho \, = \, -0.45 \pm 0.10$.

@@

The signal simulation procedure has been validated reproducing the 39.6~keV xenon line from interaction due to
$^{124}$AmBe source and has been compared to data. For the comparison the proper  $^{124}$AmBe nuclear recoil and acceptances
were simulated. The simulated events were found in agreement  with calibration data within statistical uncertainties.

@@

The main expected background contribution in the region of interest is due material radioactivity, composed mainly by photons  
interacting via compton scattering. Background contribution due to the activation of the xenon 39.6~keV line from radiogenic neutrons is expected to be negligible.
The background is modeled using data from $^{60}$Co calibration campaign, which are assumed to well represent the background density distribution 
in the cS1-cS2 plane.

@@

@@

Uncertainties on the energy scale and, more generally, related to detector response 
are parametrized using the respective uncertainties on the measure of $L_y$, $\mathcal{L}_{eff}$, $Y$, $Q_Y$ and $\rho$. The simulation shows 
that these type of uncertainties mainly affect the pdf of the signal model in the ROI, and very weakly the total signal yield. 
%For each WIMP mass we simulate several signal pseudo-sample, which are produced varying the model parameters  respectively of $\pm 1$ standard deviation. 
They are taken into account by simulating several signal pseudo-sample for each WIMP mass, the pseudo-sampleis are produced varying the model parameters respectively of $\pm 1$ standard deviation. 
For each sub-region is then computed an overall uncertainty by adding in quadrature the variations of each pseudo-sample 
%FIXME: explain this better, the up variation are separated from down variations and added togheter
with respect to nominal. Figure~\ref{fig:unc} is an example of such a systematic uncertainty computation for WIMP mass of 100~GeV.

%FIXME: express this in an understandable way
All the uncertainties discussed here are parametrized within a binned profiled likelihood function using the framework~\cite{roostat,roofit}.
All parameters related to systematic uncertainties are assumed to be normally distributed.

### 2017-01-09 Inelastic/results.tex
Some changes
first version to ranny

\section{Results}
No evidence of dark matter.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{images/data_vs_bkg.png}
  \caption{Results, comparison between data and expected background.}
  \label{fig:dataVSbkg}
\end{figure}

\begin{figure}[h!]
  \includegraphics[width=\linewidth]{images/limit_reb.png}
  \caption{Observed and expected limits.}
  \label{fig:limits}
\end{figure}

>>>>

\section{Results}
In an exposure of 34~kg of liquid xenon and 224.6~live days of data a yield of 764 events are found in the region of interest,
this is compatible with the expectation of $756 \, \pm \, 5^{(stat.)} \, \pm 55^{(syst.)}$ events from the background only hypothesis. 
Figure~\ref{fig:dataVSbkg} shows how these events are distributed in the region of interest, the bottom panel shows the ratio
between data and expected background, where the gray and orange shaded areas represent respectively statistical and systematic uncertainty 
on the background expectation.
%The distribution of events from data are compared to the background expectation in Figure~\ref{fig:dataVSbkg} and with the relative
%uncertainties.  
This result is interpreted via a binned profiled likelihood approach by means of the test statistic $\tilde{q}$
and its asymptotic distributions described in \cite{asympt}.

Assuming  an isothermal WIMP halo with a local density of $\rho_{\chi} \, = \, 0.3$~$GeV/cm^3$ , a local circular veloc-
ity of $v_0 \,= \, 220$~km/s, and a galactic escape velocity of $v_{esc} \, = \, 544$ km/s, 
\textcolor{blue}{other assumptions...,} a 90\% CL$_s$~\cite{cls} confidence level limit is  
computed on the spin dependent inelastic WIMP-nucleon cross section, $\sigma_{ine}$, as a function of the WIMP mass, $m_{\chi}$, and shown in Figure~\ref{fig:limits}.
The expecetd median sensitivity is reported with its relative one (green area) and two (yellow area) standard deviation uncertainty.
A limit is set on $\sigma_{ine}$ to $3.3 \times 10^{-38} ~cm^{2}$ for a WIMP of mass 100~GeV. This limit is compared with
\textcolor{blue}{decide which other experiment to plot.}

\newpage
\begin{figure}[t!]
  \includegraphics[width=\linewidth]{images/data_vs_bkg.png}
  \caption{Results, comparison between data and expected background.}
  \label{fig:dataVSbkg}
\end{figure}

\begin{figure}[h]
  \includegraphics[width=\linewidth]{images/limit_reb.png}
  \caption{Observed and expected limits.}
  \label{fig:limits}
\end{figure}

\newpage
